Mechanism to control synchronization in coupled area-preserving maps

### 2018-08-31 Syncarea_control/paper.tex
paper updated

\section{Delayed Synchronization Times}
\label{sec:delay}
In order to develop a numerical technique to delay synchronization, we  first 
have to identify the vicinity of regular islands in the phase space. A 
mechanism  to suppress stickiness based on the knowledge of hyperbolic and non 
hyperbolic regions in the phase has been reported recently~\cite{Kruger2015}. 
Our method, however, targets the trajectories themselves in the sticky region 
to delay the process.  It is to be noted that the proposed procedure ignores 
the details of a rather complex hierarchy of structures around regular 
islands. We employ the edge-detection algorithm due to Benkadda {\it et al.} 
\cite{Benkadda1997} to identify the edges of regular islands. For the 
numerical procedure, we divide the phase space in $100 \times 100$ grid. The 
edges of both the islands are detected by applying the standard map to point 
initiated in the chaotic sea, for $10^7$ times.  The edges thus detected in 
shown in Fig.  We have chosen the thickness parameter to be $d =0.02$ i.e. 
euclidean distance of $d$ from points on the edge indicate the extent of the 
vicinity of the island.  This vicinity, therefore, indicate the domain wherein 
the synchronization traps exist.  To demonstrate this explicitly, we compare 
the phase angles, defined by $\theta = \tan^{-1}(\frac{Q}{P}$), of the points 
on 
the numerically detected edges and the points of synchronization, as follows.

@@

The proximity parameter $d_0 = 0.02$ defines the region in the neighbourhood of the numerically detected edges wherein synchronization traps exist. The basic idea to control synchronization time is to identify the step at which the chaotic trajectory visits the numerically determined sticky neighborhood followed by a slight deflection so that the trajectory restarts at a point outside.  Our procedure depends upon how many times we deflect the trajectory  which will be referred to as step control. This means that if an $n-$step control is  employed then, the trajectory will be kicked away from the domain $n$-times during its first $n$ visits i.e. once per visit. The deflection is achieved by adding a randomly generated fraction below $0.1$ to the point visiting the domain, and therefore the maximum deflection area is roughly $1\%$ of the phase space.  This procedure applied on a given set of initial conditions of the drive and response maps, may result in four possibilities of synchronization time -- (1) successful delay  \textbf{(S)}, (2) no delay \textbf{(N)}, (3) failed to achieve synchronization \textbf{(F)}, and (4) undesirable faster synchronization \textbf{(U)}. Clearly, the only first possibility is the aim of the mechanism; the last one constitute a hazard while the other two are trivial. 
\begin{figure}[h]
	\includegraphics[scale=0.45]{S_U_Percent}
	\caption{\label{fig:Control_success}\footnotesize The curves indicate the 
		percentages of (S) successful delays shown as the blue curve, and (U) 
		undesirable fast synchronization shown as the red curve. }
\end{figure}

@@

@@

\begin{figure}[t]
	\includegraphics[scale=0.45]{Strength_con.eps}
	\caption{\label{fig:Strength_con}\footnotesize Variation of $\beta$ 
		(maximum, minimum and average) which measures the strength of successful 
		delays i.e. $\beta = \ln(\frac{\tau_d}{\tau_0}$). $\beta_{max}$ 
		indicates the maximum delay time obtained at a give n-step control 
		while $\beta_{mean}$ stands for the mean value around 2. This means 
		that on an average, synchronization times re delayed by 100 times. The 
		minimum value of $\beta$ remains 1 for the cases when no delay was 
		obtained. }
\end{figure}

The degree of delay $\beta$ may be defined as the ratio of synchronization times without numerical control, $\tau_c$ to synchronization times $\tau_0$ without control, i.e. $\beta =  \ln(\frac{\tau_d}{\tau_0}$).  Therefore, for successfully delayed synchronization $\beta>0$ indicating $\tau_c > \tau$. 
The plot in Fig.~\ref{fig:Strength_con} shows the variation of $\beta$ i.e. 
$\beta_{min}$, $\beta_{mean}$, and $\beta_{max}$  indicating minimum, mean 
and maximum $\beta$ for considered $n-$step controls. Clearly, we are able 
to attain $\beta_{max}$ values to be in the range $5$ to $14$ while mean 
values remain $\sim2$. Therefore, we are successful in achieving 
significantly large delays in synchronization times.

@@

In a given area-preserving map, for instance, the  standard map, a  
perturbation in the nonlinearity parameter $K$ in the neighborhood of a 
suitable chosen periodic point leads to generation of large coherent 
structure.   We take the same form the standard map with a modulo operation 
such that $-0.5 \leq P_n \leq 0.5, 0\leq Q_n \leq 1$ . In this form, the 
standard map is known to have a hyperbolic fixed point at 
$(0,0)$,$(0,0.5)$. We perturb the parameter $K$ to $K-\epsilon$ if $|P - P_f| 
< \delta$, $|Q-Q_f| < \delta$. For $P$ and $Q$ outside this $\delta$-strip, 
$K$ does not change.  It may be noted that the perturbed standard map remains 
to be area-preserving. The Jacobian $J$ is now given by:

@@

\vspace{0.4cm}
	where, $ \epsilon \neq 0$ if $|P_f - P_n| < \delta$, $|Q_f - Q_n| < \delta$\\ $\epsilon = 0$, otherwise.\\

\begin{figure*}[t]
	\includegraphics[scale=0.4]{Fast_sync_location.eps}
	\caption{\label{fig:location}\footnotesize The phase space of the drive 
	map at K = 6 for a couple of initial conditions where parameter 
	perturbation is applied. Synchronization occurs at the location indicate 
	by the red `plus' signs.}
\end{figure*}

@@

For the computations, we take 5 000 initial conditions chosen randomly from 
the chaotic region of the unperturbed standard map i.e. $P_0 \in \{-0.4,0.4\}$ 
and $Q_0 \in \{0.3,0.7\}$. The total number of iterations used for each 
initial condition to check for synchronization is $10^7$ and threshold of 
synchronization is $10^{-5}$ as previously.  The procedure has 
been applied to both drive and 
response map in accordance with our coupling  scheme described above. We 
choose a patch around $(P_f,Q_f)$  of length $\delta= 0.4$ which corresponds 
to about $16\%$ of the phase space area. We successfully obtain advance 
synchronization in more that $99.5\%$ of the cases among which a  couple of 
instances  has been shown in Fig.~\ref{fig:location} -- synchronization occurs 
in the neighborhood of one of the regular islands and one of their satellite 
islands.  The sign `+' in red indicate the point of synchronization. For 
details for the synchronization process, see ~\cite{Mahata2016,Das2017}.

We again define the degree of controlled synchronization as $\beta = \ln 
\frac{\tau_c}{\tau}$. For advanced synchronization, $\beta < 0$ indicating 
that synchronization times upon parameter perturbation is smaller than in 
normal course i.e. $\tau_c < \tau$. The normalized probability distribution of 
$\beta$ in shown in Fig.~\ref{fig:beta_dist}, which may be fit with a Gaussian 
of the following form:
\begin{equation}
f(\beta) =  a\exp\Big(-\frac{(\beta-\mu)^2}{2\sigma^2}\Big)
\end{equation}

@@

The parameter perturbation technique is an efficient way to decrease synchronization times with high accuracy.  The value of the perturbation imposed should be chosen suitably as a very small may not have any influence on the synchronization process. The procedure also depends on the size and location of the patch wherein the control is applied. We are yet to see if the other technique of generating coherent structures also influence synchronization in these unidirectionally coupled systems. 
\begin{figure}[b]
	%\includegraphics[scale=0.45]{sync_time_dist_beta.pdf}
    \includegraphics[scale=0.45]{Fast_strength_dist.eps}
	\caption{\label{fig:beta_dist}\footnotesize Distribution of degree of 
	ynchronization, $P(\beta)$, mean is -7.46 and standard deviation 1.67.}
\end{figure}

\section{Conclusions}
\label{sec:conclusions}
To summarize, we have developed numerical procedures to control 
synchronization in two identical standard map coupled under the drive-response 
configuration. The procedure is based on the fact that synchronization in the 
system typically occurs in the sticky neighborhoods of the regular islands in 
the phase space. We have shown that a delay in synchronization can be obtained 
by kicking the trajectories slightly away from the domain temporarily. The 
efficiency of the procedure depends upon the number of steps employed in 
control. By applying the procedure to the system on merely $1\%$ of the phase 
space area (at $K = 6$), we have successfully increased synchronization times. 
The maximum number of control steps that are used is $0.01\%$ of the total of 
number of iterations used throughout the computation.  Furthermore, the 
procedure gives variety of delays. The limitations of this procedure are, one, 
the deflection may insert the drive trajectory in the regular islands which 
would lead to undesirable faster synchronization and two, a predetermined 
synchronization time in not obtainable. Furthermore, we have demonstrated a 
simple way to fasten synchronization by the parameter perturbation technique 
which rapidly drives the chaotic trajectories in the sticky neighborhood of 
regular islands. The distribution of degree of synchronization times indicates 
the efficiency of the procedure. In both the numerical procedures, we have 
ignored the complex hierarchical structures which lead to sticky 
behavior of the chaotic orbits in the phase space. A possible future direction 
is to investigate the role of stickiness in measure synchronization in coupled 
Hamiltonian systems.

>>>>

\section{Delayed Synchronization Times}
\label{sec:delay}
In order to develop a numerical technique to delay synchronization, we  first 
have to identify the vicinity of regular islands in the phase space. A 
mechanism  to suppress stickiness based on the knowledge of hyperbolic and non 
hyperbolic regions in the phase has been reported recently~\cite{Kruger2015}. 
Our method, however, targets the trajectories themselves in the sticky region 
to delay the process.  It is to be noted that the proposed procedure ignores 
the details of a rather complex hierarchy of structures around regular 
islands. We employ the edge-detection algorithm due to Benkadda {\it et al.} 
\cite{Benkadda1997} to identify the edges of regular islands in phase space. 
For the numerical procedure, we divide the phase space in $100 \times 100$ 
grid. The edges of both the islands are detected by applying the standard map 
to point initiated in the chaotic sea, for $10^7$ times.  The edges thus 
detected in shown in Fig~\ref{fig:edge_reg_island}.  We have chosen the 
thickness parameter to be $d=0.02$ i.e. euclidean distance of $d$ from points 
on the edge indicate the extent of the vicinity of the island.  This vicinity, 
therefore, indicate the domain wherein the synchronization traps exist.  To 
demonstrate this explicitly, we compare the phase angles, defined by $\theta = 
\tan^{-1}(\frac{Q}{P}$), of the points on the numerically detected edges and 
the points of synchronization, as 
follows.

\begin{figure}[b]
    \includegraphics[scale=0.42]{edge_reg_island.eps}
    %\includegraphics[scale=.4]{Standard_Map2.eps}\\
    \caption{\label{fig:edge_reg_island} \footnotesize Example to detect the 
    edge of a regular island in the phase space using edge detection 
    algorithm  due to Benkadda {\it et al.} \cite{Benkadda1997}. }
\end{figure}

@@

The proximity parameter $d_0 = 0.02$ defines the region in the neighborhood of 
the numerically detected edges wherein synchronization traps exist. The basic 
idea to control synchronization time is to identify the step at which the 
chaotic trajectory visits the numerically determined sticky neighborhood 
followed by a slight deflection so that the trajectory restarts at a point 
outside.  Our procedure depends upon how many times we deflect the trajectory  
which will be referred to as step control. This means that if an $n-$step 
control is  employed then, the trajectory will be kicked away from the domain 
$n$-times during its first $n$ visits i.e. once per visit. The deflection is 
achieved by adding a randomly generated fraction below $0.1$ to the point 
visiting the domain, and therefore the maximum deflection area is roughly 
$1\%$ of the phase space.  This procedure applied on a given set of initial 
conditions of the drive and response maps, may result in four possibilities of 
synchronization time -- (1) successful delay  \textbf{(S)}, (2) no delay 
\textbf{(N)}, (3) failed to achieve synchronization \textbf{(F)}, and (4) 
undesirable faster synchronization \textbf{(U)}. Clearly, the only first 
possibility is the aim of the mechanism; the last one constitute a hazard 
while the other two are trivial. 
\begin{figure}[h]
	\includegraphics[scale=0.45]{S_U_Percent}
	\caption{\label{fig:Control_success}\footnotesize The curves indicate the 
		percentages of (S) successful delays shown as the blue curve, and (U) 
		undesirable fast synchronization shown as the red curve. }
\end{figure}

@@

\begin{figure}[b]
    \includegraphics[scale=0.45]{Strength_con.eps}
    \caption{\label{fig:Strength_con}\footnotesize Variation of $\beta$ 
        (maximum, minimum and average) which measures the strength of 
        successful 
        delays i.e. $\beta = \ln(\frac{\tau_d}{\tau_0}$). $\beta_{max}$ 
        indicates the maximum delay time obtained at a give n-step control 
        while $\beta_{mean}$ stands for the mean value around 2. This means 
        that on an average, synchronization times re delayed by 100 times. The 
        minimum value of $\beta$ remains 1 for the cases when no delay was 
        obtained. }
\end{figure}

@@

The degree of controlled synchronization $\beta$ may be defined as the ratio 
of synchronization times without numerical control, $\tau_c$ to 
synchronization times $\tau_0$ without control, i.e. $\beta =  
\ln(\frac{\tau_d}{\tau_0}$).  Therefore, for successfully delayed 
synchronization $\beta>0$ indicating $\tau_c > \tau$.The plot in 
Fig.~\ref{fig:Strength_con} shows the variation of $\beta$ i.e. $\beta_{min}$, 
$\beta_{mean}$, and $\beta_{max}$  indicating minimum, mean and maximum 
$\beta$ for considered $n-$step controls. Clearly, we are able to attain 
$\beta_{max}$ values to be in the range $5$ to $14$ while mean values remain 
$\sim2$. Therefore, we are successful in achieving 
significantly large delays in synchronization times.

@@

\begin{figure*}[t]
    \includegraphics[scale=0.5]{Fast_sync_location.eps}
    \caption{\label{fig:location}\footnotesize The phase space of the drive 
        map at K = 6 for a couple of initial conditions where parameter 
        perturbation is applied. Synchronization occurs at the location 
        indicate 
        by the red `plus' signs.}
\end{figure*}
In a given area-preserving map, for instance, the  standard map, a  
perturbation in the nonlinearity parameter $K$ in the neighborhood of a 
suitable chosen periodic point leads to generation of large coherent 
structure.   We take the same form the standard map with a modulo operation 
such that $-0.5 \leq P_n \leq 0.5, 0\leq Q_n \leq 1$ . In this form, the 
standard map is known to have a hyperbolic fixed point at 
$(0,0)$,$(0,0.5)$. We perturb the parameter $K$ to $K-\epsilon$ if $|P - P_f| 
< \delta$, $|Q-Q_f| < \delta$. For $P$ and $Q$ outside this $\delta$-strip, 
$K$ does not change.  It may be noted that the perturbed standard map remains 
to be area-preserving. The Jacobian $J$ is now given by:

@@

\vspace{0.2cm}
	where, $ \epsilon \neq 0$ if $|P_f - P_n| < \delta$, $|Q_f - Q_n| < 
	\delta$\\ $\epsilon = 0$, otherwise.\\

@@

For the computations, we take 5 000 initial conditions chosen randomly from 
the chaotic region of the unperturbed standard map i.e. $P_0 \in \{-0.4,0.4\}$ 
and $Q_0 \in \{0.3,0.7\}$. The total number of iterations used for each 
initial condition to check for synchronization is $10^7$ and threshold of 
synchronization is $10^{-5}$ as previously.  The procedure has 
been applied to both drive and 
response map in accordance with our coupling  scheme described above. We 
choose a patch around $(P_f,Q_f)$  of length $\delta= 0.4$ which corresponds 
to about $16\%$ of the phase space area. We successfully obtain advance 
synchronization in more than $99.5\%$ of the cases among which a  couple of 
instances  has been shown in Fig.~\ref{fig:location} -- synchronization occurs 
in the neighborhood of one of the regular islands and one of their satellite 
islands.  The sign `+' in red indicate the point of synchronization. For 
details for the synchronization process, see ~\cite{Mahata2016,Das2017}.
\begin{figure}[h]
    \includegraphics[scale=0.8]{sync_time_beta.eps}
    \caption{\label{fig:beta_dist}\footnotesize Distribution of degree of 
        controlled synchronization, $P(\beta)$, mean is -7.46 and standard 
        deviation is 1.67.}
\end{figure}
We again define the degree of controlled synchronization as $\beta = \ln 
(\frac{\tau_c}{\tau}$). For advanced synchronization, $\beta < 0$ indicating 
that synchronization times upon parameter perturbation is smaller than in 
normal course i.e. $\tau_c < \tau$. The normalized probability distribution of 
$\beta$ in shown in Fig.~\ref{fig:beta_dist}, which may be fit with a Gaussian 
of the following form:
\begin{equation}
f(\beta) =  a\exp\Big(-\frac{(\beta-\mu)^2}{2\sigma^2}\Big)
\end{equation}

@@

The parameter perturbation technique is an efficient way to decrease synchronization times with high accuracy.  The value of the perturbation imposed should be chosen suitably as a very small may not have any influence on the synchronization process. The procedure also depends on the size and location of the patch wherein the control is applied. We are yet to see if the other technique of generating coherent structures also influence synchronization in these unidirectionally coupled systems.

\section{Conclusions}
\label{sec:conclusions}
To summarize, we have developed numerical procedures to control 
synchronization in two identical standard map coupled under the drive-response 
configuration. The procedure is based on the fact that synchronization in the 
system typically occurs in the sticky neighborhoods of the regular islands in 
the phase space. We have shown that a delay in synchronization can be obtained 
by kicking the trajectories slightly away from the domain temporarily. The 
efficiency of the procedure depends upon the number of steps employed in 
control. By applying the procedure to the system on merely $1\%$ of the phase 
space area (at $K = 6$), we have successfully increased synchronization times. 
The maximum number of control steps that are used is $0.01\%$ of the total of 
number of iterations used throughout the computation.  Furthermore, the 
procedure gives variety of delays. The limitations of this procedure are, one, 
the deflection may insert the drive trajectory in the regular islands which 
would lead to undesirable faster synchronization and two, a predetermined 
synchronization time in not obtainable. Furthermore, we have demonstrated a 
simple way to fasten synchronization by the parameter perturbation technique 
which rapidly drives the chaotic trajectories in the sticky neighborhood of 
regular islands. The distribution of degree of controlled synchronization 
times indicates the efficiency of the procedure. In both the numerical 
procedures, we have ignored the complex hierarchical structures which lead to 
sticky behavior of the chaotic orbits in the phase space. A possible future 
direction is to investigate the role of stickiness in measure synchronization 
in coupled Hamiltonian systems.
